---
title: "Last.fm Recommender"
author: "Brian"
date: "September 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1) SQL Clickstream Query (MySQL)
Find the count of users who install the app (i.e. with FIRST_INSTALL event) on 2017-04-01 and use our app at least once (i.e. with any event) between 2017-04-02 and 2017-04-08.
Table name: piwik_track
- Table fields:
- uid [string]: Unique ID of user
- time [datetime]: The time when we receive the event
- event_name string ENUM { FIRST_INSTALL, ... } type of event
```{r Clickstream Query SQL}

#SELECT COUNT(DISTINCT uid) AS usercount
#FROM piwik_track
#WHERE 
#  (DATE(time) = '2017-04-01' AND event_name LIKE '%FIRST_INSTALL%') AND
#  (DATE(time) BETWEEN '2017-04-02' AND '2017-04-08')

SELECT COUNT(DISTINCT uid) AS usercount
FROM (
      SELECT uid, time, event_name
        FROM piwik_track
        WHERE
            (DATE(time) BETWEEN '2017-04-02' AND '2017-04-08') AND
            (event_name NOT LIKE '%FIRST_INSTALL%')
    )
WHERE
    (DATE(time) = '2017-04-01' AND event_name LIKE '%FIRST_INSTALL%') 
```

# Libraries for R
```{r}
library(dplyr)
library(lubridate)
library(tidyverse)
library(knitr)
library(wordcloud)
```


# 2) Raw Data Analytics (R)
Count the total data transfer (namely the sum of the file size of individual request) caused by JPEG files (i.e. URI ends with jpg) from 24th Aug to 25th Aug
```{r Raw Data Analytics}


df <- read.table("./logfile.txt", sep="\t", header=TRUE, stringsAsFactors = FALSE)
df$date <- ymd(df$date)

df %>%  filter(date >= "2017-08-24" & date <= "2017-08-25") %>% filter(grepl(".jpg$",url)) %>% select(size) %>% sum()
# [1] 85251
```

# 3) Last.fm Recommendation Engine (R)
```{r pressure, echo=FALSE}
#train <- read.csv("./offsite-tagging-training-set.csv", #stringsAsFactors = FALSE, header = TRUE, sep=",", encoding = "UTF-8")

# Read profile and plays TSV data

profile <- read.table("./lastfm-dataset-360K/usersha1-profile.tsv", sep="\t", header=FALSE, stringsAsFactors = FALSE)
colnames(profile) <- c("userid","gender","age","country","signup")

playcount <- read.table("./lastfm-dataset-360K/usersha1-artmbid-artname-plays.tsv", sep="\t", fill=TRUE, quote="", stringsAsFactors = FALSE)
colnames(playcount) <- c("userid","artistid","artist","plays")

```

Wordcloud visualization for male and female
```{r}
x <- merge(playcount, profile, all.x=TRUE, by.x="userid", by.y="userid")

female <- x %>% filter(gender=="f")

female.aggr <- aggregate(plays ~ artist, female, sum)
female.aggr <- female.aggr[order(-female.aggr$plays),]
female.aggr[,"female_rank"] <- rank(-female.aggr$plays, ties.method="first")

male <- x %>% filter(gender=="m")

male.aggr <- aggregate(plays ~ artist, male, sum)
male.aggr <- male.aggr[order(-male.aggr$plays),]
male.aggr[,"male_rank"] <- rank(-male.aggr$plays, ties.method="first")


plot.female <- wordcloud(female.aggr$artist, female.aggr$plays, max.words=100)
plot.male <- wordcloud(male.aggr$artist, male.aggr$plays, max.words=100)

z <- merge(male.aggr, female.aggr, all.x=TRUE, all.y=TRUE, by.x="artist", by.y="artist")
z <- z[order(-z$plays.x),]
kable(head(z, 50))
# As seen in the Top 50 artists, males and females have different tastes in music
```

# Attempt at Recommender with raw data for female users
```{r}
library(recommenderlab)
library(reshape2)

# Looking at number of female users and how many artists they listen to
length(unique(female$userid))
# [1] 1328 unique female users
length(unique(female$artistid))
# [1] 15091 unique artists

# No. of rows with empty artistid, and remove them
nrow(female[female$artistid == "",])
# [1] 837
female <- female %>% filter(artistid != "")

# No. of rows with empty artist, and remove them
nrow(female[female$artist == "",])
# [1] 5
female <- female %>% filter(artist != "")

# Transform data to user/artist matrix
female.matrix <- acast(female, userid ~ artistid, fun.aggregate=sum, value.var="plays")
# Double check if matrix has correct dimensions
nrow(female.matrix)
# [1] 1328 = total unique users
ncol(female.matrix)
# [1] 15089 = total unique artists

sum(female$plays, na.rm=TRUE)
# [1] 12783183 = total plays in dataframe
sum(female.matrix, na.rm=TRUE)
# [1] 12783183 = total plays in matrix

# Transform into realRatingMatrix for recommenderlab
female.real.matrix <- as(female.matrix,"realRatingMatrix")

# Normalize plays
female.real.matrix2 <- normalize(female.real.matrix, method="Z-score")

rec.model <-Recommender(female.real.matrix, method = "UBCF", param=list(normalize = "Z-score",method="Cosine",nn=5, minRating=1))

as(predict(rec.model, female.real.matrix["00000c289a1829a808ac09c00daf10bc3c4e223b",], n=5), "list")
```

```{r}
library(recommenderlab)
library(reshape2)

# Looking at number of plays users and how many artists they listen to
length(unique(playcount$userid))
# [1] 8875 unique plays users
length(unique(playcount$artistid))
# [1] 46551 unique artists

# No. of rows with empty artistid, and remove them
nrow(playcount[playcount$artistid == "",])
# [1] 837
playcount <- playcount %>% filter(artistid != "")

# No. of rows with empty artist, and remove them
nrow(playcount[playcount$artist == "",])
# [1] 5
playcount <- playcount %>% filter(artist != "")

# Transform data to user/artist matrix
playcount.matrix <- acast(playcount, userid ~ artistid, fun.aggregate=sum, value.var="plays")
# Double check if matrix has correct dimensions
nrow(playcount.matrix)
# [1] 1328 = total unique users
ncol(playcount.matrix)
# [1] 15089 = total unique artists

sum(playcount$plays, na.rm=TRUE)
# [1] 12783183 = total plays in dataframe
sum(playcount.matrix, na.rm=TRUE)
# [1] 12783183 = total plays in matrix

# Transform into realRatingMatrix for recommenderlab
playcount.real.matrix<- as(playcount.matrix,"realRatingMatrix")

rec.model <-Recommender(playcount.real.matrix, method = "UBCF")

predicted <- predict(rec.model, playcount.real.matrix[1,], n=5)
as(predicted, "list")
```
```

```{r}
z <- aggregate(plays ~ userid, playcount, sum)
z <- z[order(-plays),]

z1 <- head(z, 500)
z2 <- as.vector(z1$userid)

```

