---
title: "Last.fm Recommender"
author: "Brian"
date: "September 28, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1) SQL Clickstream Query (MySQL)
Find the count of users who install the app (i.e. with FIRST_INSTALL event) on 2017-04-01 and use our app at least once (i.e. with any event) between 2017-04-02 and 2017-04-08.
Table name: piwik_track
- Table fields:
- uid [string]: Unique ID of user
- time [datetime]: The time when we receive the event
- event_name string ENUM { FIRST_INSTALL, ... } type of event

The database used for this query was MySQL
```{r Clickstream Query MySQL}

# CREATE DATABASE testdb;
# USE testdb;

#create table piwik_track
#(
#time datetime,
#uid varchar(256),
#event_name varchar(256),
#source_ip varchar(50),
#);

SELECT COUNT(DISTINCT uid)
      AS usercount
      FROM piwik_track
      WHERE 
	       	  (uid IN
			          (SELECT uid
			           FROM piwik_track
			           WHERE (DATE(time) = '2017-04-01' AND event_name LIKE '%FIRST_INSTALL%')
		           	)
	         	) AND
	        	(DATE(time) BETWEEN '2017-04-02' AND '2017-04-08') AND
	        	(event_name NOT LIKE '%FIRST_INSTALL%')
;
      
# usercount
# 5
```

# Libraries for R

These libraries are required for parts 2 and 3 (only dplyr and lubridate are required for Part 2)
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(tidyverse)
library(knitr)
library(wordcloud)
library(recommenderlab)
library(reshape2)
library(shiny)
library(RColorBrewer)
```


## 2) Raw Data Analytics (R)
Count the total data transfer (namely the sum of the file size of individual request) caused by JPEG files (i.e. URI ends with jpg) from 24th Aug to 25th Aug

R was used for this example, and the data from the PDF was used for "logfile.txt"
```{r Raw Data Analytics}
# library(dplyr)
# library(lubridate)

df <- read.table("./logfile.txt", sep="\t", header=TRUE, stringsAsFactors = FALSE)
df$date <- ymd(df$date)

df %>%  filter(date >= "2017-08-24" & date <= "2017-08-25") %>% filter(grepl(".jpg$",url)) %>% select(size) %>% sum()
# [1] 85251
```

## 3) Last.fm Recommendation Engine (R)

A recommendation engine was created in R for the Last.fm 360k dataset, featuring users profiles and playcounts of artists for those users. 

The goal of this recommendation engine is to recommend similar artists to users, based on their existing artist preferences - this is estimated by observing the number of plays each user has for each artist. Only the female users were considered for this recommendation engine to limit the scope of the engine, and because the listening preferences of males and females differ significantly in this dataset.

# Environment Setup
To run this recommendation engine code, please run all chunks below in the current file "lastfm-recommender.Rmd" in RStudio, cloned from the Github repo. The last.fm data file (lastfm-dataset-360K/usersha1-profile.tsv, lastfm-dataset-360K/usersha1-artmbid-artname-plays.tsv) should be placed in the root of the repo. Additionally, the R packages above should be installed to execute the R script.

This executes the recommendation engine (User Based Collaborative Filtering), and then run the Shiny app.R found in the "user_recommender" folder for a rough prototype of the UI of the recommender.

# Loading data from Last.fm dataset
Load TSV files from http://www.dtic.upf.edu/~ocelma/MusicRecommendationDataset/lastfm-360K.html, placed in the root of this git project
```{r pressure, echo=FALSE}

# Read profile and playcount TSV data
profile <- read.table("./lastfm-dataset-360K/usersha1-profile.tsv", sep="\t", header=FALSE, stringsAsFactors = FALSE)
colnames(profile) <- c("userid","gender","age","country","signup")

playcount <- read.table("./lastfm-dataset-360K/usersha1-artmbid-artname-plays.tsv", sep="\t", fill=TRUE, quote="", stringsAsFactors = FALSE)
colnames(playcount) <- c("userid","artistid","artist","plays")

## Merge profile and playcount
full <- merge(playcount, profile, all.x=TRUE, by.x="userid", by.y="userid")

## Create Male/Female datasets from playcount
female <- full %>% filter(gender=="f")
male <- full %>% filter(gender=="m")

## Create dimension table for unique artists and unique users - these will be used for the Shiny UI
unique_artists <- full %>% select(artistid, artist) %>% unique() %>% as.data.frame() %>% filter(artistid != "")
unique_users <- unique(female$userid)

# Preview of the important tables from this chunk
glimpse(female)
glimpse(unique_artists)
glimpse(unique_users)
```

# Wordcloud visualization and simple EDA for male and female

The goal of this section is to create some basic visualizations to view the most played artists for males and females. This is used to visualize whether males and females have similar music preferences, and whether the recommendation engine should consider all users, or should view males and females as separate subsets with different preferences.

The most popular artists by playcount are aggregated and ranked, and are plot as bar charts and wordclouds as a simple EDA exercise.
```{r}
# Aggregation of top artists for female users
# Ranking for artists is determined using rank(ties.method="first"), such that a tie in ranking results in both artist having the same lower rank - i.e. 1, 2,2,4,5)
female.aggr <- aggregate(plays ~ artist, female, sum)
female.aggr <- female.aggr[order(-female.aggr$plays),]
female.aggr[,"female_rank"] <- rank(-female.aggr$plays, ties.method="min")

# Aggregation of top artists for male users
male.aggr <- aggregate(plays ~ artist, male, sum)
male.aggr <- male.aggr[order(-male.aggr$plays),]
male.aggr[,"male_rank"] <- rank(-male.aggr$plays, ties.method="min")

# Wordclouds for males and females for top 30 artists
color1 <- brewer.pal(8,"Dark2")
plot.female <- wordcloud(female.aggr$artist, female.aggr$plays, max.words=30, scale=c(2,.1), colors=color1)
plot.male <- wordcloud(male.aggr$artist, male.aggr$plays, max.words=30, scale=c(2,.1), colors=color1)

# Show top 30 rankings for males and females merged together, showing rank difference for females vs males in terms of artist plays
full.aggr <- merge(male.aggr, female.aggr, all.x=TRUE, all.y=TRUE, by.x="artist", by.y="artist")
full.aggr <- full.aggr[order(-full.aggr$plays.y),]
full.aggr$rank_diff <- full.aggr$female_rank - full.aggr$male_rank
kable(head(full.aggr, 30))

# Bar plots showing Top 30 artists for Male and Female users
barcolor = brewer.pal(3,"Blues")
ggplot(head(full.aggr[,c("artist","plays.x")],30), aes(x = reorder(artist,plays.x), y=plays.x)) + geom_bar(stat="identity", fill="lightskyblue") + coord_flip() + ggtitle("Top Artists for Male Last.fm Users")
ggplot(head(full.aggr[,c("artist","plays.y")],30), aes(x = reorder(artist,plays.y), y=plays.y)) + geom_bar(stat="identity", fill="violetred") + coord_flip() + ggtitle("Top Artists for Female Last.fm Users")

# Calculating the rank difference for the top 50 female artists. The value represents an approximation of how popuar the artist is for females compared to males
mean(head(full.aggr$rank_diff,50), na.rm=TRUE)
# [1] -246.06
```

# User Based Collaborative Filter to create Recommender for female users (using only female users)
```{r}
# Looking at number of female users and how many artists they listen to
length(unique(female$userid))
# [1] 1328 unique female users
length(unique(female$artistid))
# [1] 15091 unique artists

# No. of rows with empty artistid, and remove them
nrow(female[female$artistid == "",])
# [1] 837
female <- female %>% filter(artistid != "")

# No. of rows with empty artist, and remove them
nrow(female[female$artist == "",])
# [1] 5
female <- female %>% filter(artist != "")

# Transform data to user/artist matrix
female.matrix <- acast(female, userid ~ artistid, fun.aggregate=sum, value.var="plays")
# Double check if matrix has correct dimensions
nrow(female.matrix)
# [1] 1328 = total unique users
ncol(female.matrix)
# [1] 15089 = total unique artists

sum(female$plays, na.rm=TRUE)
# [1] 12783183 = total plays in dataframe
sum(female.matrix, na.rm=TRUE)
# [1] 12783183 = total plays in matrix

# Convert all 0 to NA for recommenderlab to work
female.matrix[female.matrix == 0] <- NA

# Transform into realRatingMatrix for recommenderlab
female.real <- as(female.matrix,"realRatingMatrix")

# Used first 929 female users as training data (split of 70/30)
# Method for normalization was centering instead of Z-score, may yield improvements with Z-score or other normalization
# nn = 20 nearest neighbours can be hypertuned using cross-validation for better model in the future
rec.model <-Recommender(female.real[1:929], method = "UBCF", param=list(normalize = "center",method="Cosine",nn=20))
 
# Used remaining users 930-1328 as test data
as(predict(rec.model, female.real[930], n=10), "list")
# Recommendations can be done for users 930-1328
```

# Shiny App
After running "rec.model", the Shiny App located in ./user_recommender/app.R can be run. The app is a mock-up of a user's Last.fm profile, featuring their top played artists (by play count), as well as a list of top 10 recommended artists based on the rec.model (UBCF). The Shiny app will generate top 10 played artists and top 10 recommended artists for the selected user.

# System Design Limitations
The system is based on a database of user profiles and artist play counts. As there is limited detail on the time scale (when each artist was played, time series trends on when each artist was played), as well as no information on number of songs/which songs played, the level of recommendation is limited by the current dataset.

# Future Enhancements 
For a more enhanced model, individual songs can be recommended as well. However, the UBCF model will be much larger/more expensive due to the number of songs available per artist. For this model, it is recommended to use the "POPULAR" setting to only recommend the most popular songs for each artist (based on number of plays per song)

# Concerns if system is put into production
- Due to the large number of different artists on Last.fm, including genres, languages, types of music (vocal, instrumental, classical, soundtrack, or non-music such as podcasts, comedy albums), it becomes hard to recommend items based solely on UBCF or IBCF. There needs to be a more robust and less expensive method than to create a massive matrix. It is suggested to cluster and segment users based on their age, country, genres and other demographics/behavior to create micro-segments and run UBCF on the micro-clusters.

- One of the challenges is to verify the accuracy of the recommendations made by the user. There needed to be feedback at production to indicate whether the user selects the recommended artists (clickthrough/artist page views?), as well as number of plays for each of the recommended artists. The best tracking is if a system that tracks skipped songs/songs that are not played to x% (e.g. 60%) completion, suggesting the user does not like the recommended song/artist. These systems are implemented by YouTube, Spotify, etc. to gauge user interest in the song/video.

- There should be enhancements to the system that take into account the recency of the plays/artists (Last.fm already implements this with its "Top Played This Month", "Top Played Today" etc.) in addition to the frequency of plays. This makes it easier for the system to recommend songs/artists for the user, as user tastes/interest change quickly. However, it is a challenge to not overfit for this data and only make the recommendations based on the most recent plays. Spotify's algorithm runs into this issue with its "Discover Weekly" playlist recommending songs based on recent plays - it overfits and will recommend entire playlists of soundtrack songs if the user recently plays many soundtrack songs (even if the songs are not played in their entirety, or the same soundtrack(s) are looped). Therefore, the system should take into account genre or language or other factors. Ideally, the user would receive multiple lists of recommendations based on set macro-genres (e.g. "I want recommendations for Pop, based on my history" or "I want recommendations for Mandarin songs based on my play history"), or the user would be able to subscribe to certain genres to reduce the computational cost for each user's lists.




## Additional tested content for Question 3

# Seeing if PCA can be used to reduce dimensionality
Attempted to use Principal Component Analysis to reduce number of different artists into groups, but was not very effective overall.

Based on the plot, 10 PCs only explained about 60% of the total variance in the dataset, indicating that artists were not 

As there were no genre or other artist tags within the Last.fm data, we are unable to deduce whether these PCs are a representation of the Genres or other categories of artists, or there are other segments of artists that can be used for better recommendations in the future.

A better use would be to utilize artist's "Genre", "Decade" (e.g. 60s, 70s) or other tags available from MusicBrains/Last.fm as features to improve the recommendation engine instead of PCA.
```{r}
# Change NAs to 0 in matrix
female.matrix[is.na(female.matrix)] <- 0

# Find and remove zero variance columns, as PCA does not work if columns are entirely zero variance
which(apply(female.matrix, 2, var)==0)
female.matrix <- female.matrix[ , apply(female.matrix, 2, var) != 0]

# Calculate PCA with normalization and centered to 0
female.pca <- prcomp(female.matrix, center = TRUE, scale = TRUE)
plot(female.pca, type="l")

# Biplot to view direction of principal components
# Biplot was attempted but it does not yield much information about the PCs that is useful at this point
#biplot(female.pca, scale = 0)

```

# Item based Collaborative Filtering
An attempt was made to use Item Based Collaborative Filtering for a recommendation engine. However, performance was much worse without significant improvement to the recommendations, due to the size of the matrix required for cosine similarity between 15,000+ artists.

The top 100 artists for females was used in this version below (100x100 matrix for cosine similarity) as a proof of concept. However, User-based seems to be more effective.
```{r}
#library(lsa)
# Library for cosine similarity calculation, was not used for the final IBCF

# Try top 100 female artists only for performance
top100 <- female.aggr %>% filter(female_rank <= 100) %>% select(artistid)
top100 <- as.vector(top100[,1])

# Keep only rows with artistid in top100
femaletop100 <- female %>% filter(artistid %in% top100)
femaletop100 <- femaletop100 %>% filter(artistid != "")

#
femaletop100.matrix <- acast(femaletop100, userid ~ artistid, fun.aggregate=sum, value.var="plays")

# Convert NAs to zero
femaletop100.matrix[femaletop100.matrix == 0] <- NA

# Artist Cosine Similarity (Item-based); cosine function from lsa package
#artist_sim <- cosine(femaletop100.matrix)

femaletop100.real <- as(femaletop100.matrix,"realRatingMatrix")

rec.model <-Recommender(femaletop100.real[1:1000], method = "IBCF")

predict_user <- predict(rec.model, femaletop100.real[1001:nrow(femaletop100.matrix)], n=5)

as(predict_user,"list")
```

